---
title: "Integrating genomic and tagging data reveals spatio-temporal population structure in Northeast Atlantic European sea bass"
author: "Gagnaire et al. (2026)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 7, dpi = 300)
```

# Required packages

```{r packages}
library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)
library(tibble)
library(patchwork)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(RColorBrewer)
library(adegenet)
library(hierfstat)
library(vegan)
library(ggrepel)
library(viridis)
library(cowplot)
library(scales)
library(reshape2)
```

# Data loading

```{r data-loading}
# Genomic data (SNP genotype matrix: individuals x SNPs)
geno <- read.table("data/genotypes.txt", header = TRUE, row.names = 1)

# Population metadata (individual ID, sampling location, year, coordinates)
metadata <- read.csv("data/metadata.csv")

# Tagging data (individual ID, release location, recapture location, date)
tagging <- read.csv("data/tagging_data.csv")

# FST matrix between populations
fst_matrix <- read.csv("data/fst_matrix.csv", row.names = 1)

# Admixture results (Q-matrix from ADMIXTURE or equivalent)
admix_K2 <- read.table("data/admixture_K2.txt")
admix_K3 <- read.table("data/admixture_K3.txt")

# Genetic diversity summary per population
div_stats <- read.csv("data/diversity_statistics.csv")
```

---

## FIGURE 1 ##

Sampling map showing the collection sites of European sea bass across the Northeast Atlantic
and the Mediterranean Sea, including sample sizes and the geographic context.

```{r figure1, fig.cap="Figure 1. Sampling map of European sea bass."}

# Load coastline and country boundaries
world <- ne_countries(scale = "medium", returnclass = "sf")
europe <- world[world$continent == "Europe" | world$admin %in%
                  c("Morocco", "Algeria", "Tunisia", "Libya"), ]

# Define color palette for sampling regions
region_colors <- c(
  "Bay of Biscay"        = "#2166AC",
  "English Channel"      = "#4DAC26",
  "North Sea"            = "#1A9850",
  "Iberian Atlantic"     = "#D73027",
  "Western Mediterranean" = "#F46D43",
  "Eastern Mediterranean" = "#FDAE61"
)

# Compute sample size per site for point scaling
site_summary <- metadata %>%
  group_by(site, longitude, latitude, region) %>%
  summarise(n = n(), .groups = "drop")

# Main map
fig1_map <- ggplot() +
  geom_sf(data = europe, fill = "grey85", color = "grey60", linewidth = 0.3) +
  geom_point(data = site_summary,
             aes(x = longitude, y = latitude,
                 fill = region, size = n),
             shape = 21, color = "white", stroke = 0.5, alpha = 0.9) +
  geom_text_repel(data = site_summary,
                  aes(x = longitude, y = latitude, label = site),
                  size = 2.8, box.padding = 0.4, point.padding = 0.3,
                  segment.color = "grey50", segment.size = 0.3) +
  scale_fill_manual(values = region_colors, name = "Region") +
  scale_size_continuous(name = "Sample size", range = c(2, 8),
                        breaks = c(10, 30, 60, 100)) +
  coord_sf(xlim = c(-20, 40), ylim = c(28, 62), expand = FALSE) +
  labs(x = "Longitude", y = "Latitude") +
  theme_bw(base_size = 12) +
  theme(
    legend.position  = "right",
    panel.background = element_rect(fill = "#D6EAF8"),
    panel.grid.major = element_line(color = "white", linewidth = 0.3)
  )

# Inset bar chart of sample sizes per region
fig1_inset <- ggplot(site_summary %>%
                       group_by(region) %>%
                       summarise(total = sum(n), .groups = "drop"),
                     aes(x = reorder(region, -total), y = total, fill = region)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = region_colors) +
  labs(x = NULL, y = "N individuals") +
  theme_classic(base_size = 9) +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))

# Combine map and inset
figure1 <- ggdraw(fig1_map) +
  draw_plot(fig1_inset, x = 0.60, y = 0.05, width = 0.38, height = 0.30)

figure1

ggsave("figures/Figure1_sampling_map.pdf", figure1,
       width = 10, height = 7, useDingbats = FALSE)
ggsave("figures/Figure1_sampling_map.png", figure1,
       width = 10, height = 7, dpi = 300)
```

---

## FIGURE 2 ##

Principal Component Analysis (PCA) of genome-wide SNP data illustrating the main axes
of genomic variation among individuals and sampling regions.

```{r figure2, fig.cap="Figure 2. PCA of genome-wide SNP variation."}

# Perform PCA on the genotype matrix (coded as allele dosage 0/1/2)
# Replace missing values with column means before PCA
geno_imp <- apply(geno, 2, function(x) {
  x[is.na(x)] <- mean(x, na.rm = TRUE)
  x
})
pca_res <- prcomp(geno_imp, scale. = TRUE)

# Proportion of variance explained
var_exp <- round(100 * pca_res$sdev^2 / sum(pca_res$sdev^2), 2)

# Combine PCA scores with metadata
pca_df <- data.frame(
  PC1    = pca_res$x[, 1],
  PC2    = pca_res$x[, 2],
  PC3    = pca_res$x[, 3],
  PC4    = pca_res$x[, 4],
  ID     = rownames(pca_res$x)
) %>%
  left_join(metadata, by = c("ID" = "individual_id"))

# PC1 vs PC2
p_12 <- ggplot(pca_df, aes(x = PC1, y = PC2, color = region, shape = region)) +
  geom_point(alpha = 0.75, size = 2.5) +
  stat_ellipse(aes(group = region), type = "t", level = 0.9,
               linewidth = 0.6, linetype = "dashed") +
  scale_color_manual(values = region_colors, name = "Region") +
  scale_shape_manual(values = c(16, 17, 15, 18, 8, 7), name = "Region") +
  labs(x = paste0("PC1 (", var_exp[1], "%)"),
       y = paste0("PC2 (", var_exp[2], "%)")) +
  theme_bw(base_size = 12) +
  theme(legend.position = "none")

# PC3 vs PC4
p_34 <- ggplot(pca_df, aes(x = PC3, y = PC4, color = region, shape = region)) +
  geom_point(alpha = 0.75, size = 2.5) +
  stat_ellipse(aes(group = region), type = "t", level = 0.9,
               linewidth = 0.6, linetype = "dashed") +
  scale_color_manual(values = region_colors, name = "Region") +
  scale_shape_manual(values = c(16, 17, 15, 18, 8, 7), name = "Region") +
  labs(x = paste0("PC3 (", var_exp[3], "%)"),
       y = paste0("PC4 (", var_exp[4], "%)")) +
  theme_bw(base_size = 12)

# Scree plot (variance explained per PC)
p_scree <- data.frame(PC = 1:20, VarExp = var_exp[1:20]) %>%
  ggplot(aes(x = PC, y = VarExp)) +
  geom_col(fill = "steelblue", width = 0.7) +
  geom_line(color = "darkred", linewidth = 0.7) +
  geom_point(color = "darkred", size = 2) +
  labs(x = "Principal Component", y = "Variance explained (%)") +
  theme_classic(base_size = 10)

figure2 <- (p_12 | p_34) / p_scree +
  plot_layout(heights = c(3, 1)) +
  plot_annotation(tag_levels = "A")

figure2

ggsave("figures/Figure2_PCA.pdf", figure2,
       width = 12, height = 8, useDingbats = FALSE)
ggsave("figures/Figure2_PCA.png", figure2,
       width = 12, height = 8, dpi = 300)
```

---

## FIGURE 3 ##

Ancestry proportions estimated by ADMIXTURE for K = 2 and K = 3 genetic clusters,
sorted geographically along a west-to-east axis.

```{r figure3, fig.cap="Figure 3. Individual ancestry proportions from ADMIXTURE."}

# Attach metadata to admixture results (individuals must be in the same order)
admix_K2_df <- cbind(metadata, admix_K2)
admix_K3_df <- cbind(metadata, admix_K3)

# Geographic order: sort populations from west (Atlantic) to east (Mediterranean)
geo_order <- metadata %>%
  group_by(site, region) %>%
  summarise(mean_lon = mean(longitude), .groups = "drop") %>%
  arrange(mean_lon) %>%
  pull(site)

# Helper function to reshape and plot admixture bar charts
plot_admixture <- function(df, K, colors, title) {
  q_cols <- paste0("V", seq_len(K))
  df_long <- df %>%
    mutate(individual_id = factor(individual_id,
                                  levels = df$individual_id[order(match(df$site, geo_order),
                                                                  df[[q_cols[1]]])]),
           site = factor(site, levels = geo_order)) %>%
    pivot_longer(cols = all_of(q_cols),
                 names_to  = "cluster",
                 values_to = "ancestry")

  ggplot(df_long, aes(x = individual_id, y = ancestry, fill = cluster)) +
    geom_col(width = 1, position = "stack") +
    scale_fill_manual(values = colors, name = "Cluster") +
    facet_grid(~ site, scales = "free_x", space = "free_x") +
    labs(x = NULL, y = "Ancestry proportion", title = title) +
    theme_minimal(base_size = 10) +
    theme(
      axis.text.x      = element_blank(),
      axis.ticks.x     = element_blank(),
      strip.text       = element_text(angle = 45, hjust = 1, size = 8),
      panel.spacing    = unit(0.05, "lines"),
      legend.position  = "bottom"
    )
}

p_K2 <- plot_admixture(admix_K2_df, K = 2,
                       colors = c("#2166AC", "#D73027"), title = "K = 2")
p_K3 <- plot_admixture(admix_K3_df, K = 3,
                       colors = c("#2166AC", "#4DAC26", "#D73027"), title = "K = 3")

figure3 <- p_K2 / p_K3 +
  plot_annotation(tag_levels = "A")

figure3

ggsave("figures/Figure3_admixture.pdf", figure3,
       width = 14, height = 6, useDingbats = FALSE)
ggsave("figures/Figure3_admixture.png", figure3,
       width = 14, height = 6, dpi = 300)
```

---

## FIGURE 4 ##

Pairwise FST values between sampling sites and an Isolation-by-Distance (IBD) test
relating genetic differentiation to geographic distance.

```{r figure4, fig.cap="Figure 4. Pairwise FST heatmap and Isolation-by-Distance."}

# --- FST heatmap ---
fst_long <- fst_matrix %>%
  rownames_to_column("Pop1") %>%
  pivot_longer(-Pop1, names_to = "Pop2", values_to = "Fst") %>%
  mutate(Fst = ifelse(is.na(Fst) | Pop1 == Pop2, NA_real_, Fst))

# Order populations geographically (west to east)
pop_order <- geo_order[geo_order %in% rownames(fst_matrix)]
fst_long$Pop1 <- factor(fst_long$Pop1, levels = pop_order)
fst_long$Pop2 <- factor(fst_long$Pop2, levels = pop_order)

p_fst_heat <- ggplot(fst_long, aes(x = Pop2, y = Pop1, fill = Fst)) +
  geom_tile(color = "white", linewidth = 0.4) +
  geom_text(aes(label = ifelse(!is.na(Fst), sprintf("%.3f", Fst), "")),
            size = 2.8) +
  scale_fill_gradient2(low = "white", mid = "#FEE090", high = "#D73027",
                       midpoint = median(fst_long$Fst, na.rm = TRUE),
                       na.value = "grey95", name = expression(italic(F)[ST])) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x   = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )

# --- Isolation by Distance ---
# Compute mean coordinates per population
pop_coords <- metadata %>%
  group_by(site) %>%
  summarise(lon = mean(longitude), lat = mean(latitude), .groups = "drop") %>%
  filter(site %in% pop_order)

# Geographic distance matrix (great-circle, in km)
coord_mat    <- as.matrix(pop_coords[, c("lon", "lat")])
rownames(coord_mat) <- pop_coords$site
geo_dist     <- as.dist(fields::rdist.earth(coord_mat, miles = FALSE))

# FST / (1 - FST) linearization
fst_sym      <- as.matrix(fst_matrix)[pop_order, pop_order]
fst_lin      <- as.dist(fst_sym / (1 - fst_sym))

# Mantel test
mantel_res   <- vegan::mantel(fst_lin, geo_dist, method = "pearson",
                               permutations = 999)

# Data frame for scatter plot
ibd_df <- data.frame(
  geo_dist = as.vector(geo_dist) / 1000,   # km -> 1000 km
  fst_lin  = as.vector(fst_lin)
)

p_ibd <- ggplot(ibd_df, aes(x = geo_dist, y = fst_lin)) +
  geom_point(alpha = 0.6, color = "#2166AC", size = 2.5) +
  geom_smooth(method = "lm", color = "#D73027", fill = "#FDAE61",
              linewidth = 1, se = TRUE) +
  annotate("text", x = Inf, y = Inf,
           label = sprintf("Mantel r = %.3f\np = %.3f",
                           mantel_res$statistic, mantel_res$signif),
           hjust = 1.1, vjust = 1.5, size = 4) +
  labs(x = "Geographic distance (× 1,000 km)",
       y = expression(italic(F)[ST] / (1 - italic(F)[ST]))) +
  theme_bw(base_size = 12)

figure4 <- p_fst_heat | p_ibd +
  plot_annotation(tag_levels = "A")

figure4

ggsave("figures/Figure4_FST_IBD.pdf", figure4,
       width = 14, height = 6, useDingbats = FALSE)
ggsave("figures/Figure4_FST_IBD.png", figure4,
       width = 14, height = 6, dpi = 300)
```

---

## FIGURE 5 ##

Electronic tagging results showing the displacement trajectories of individual sea bass
and the seasonal distribution of movements across the Northeast Atlantic.

```{r figure5, fig.cap="Figure 5. Individual displacement and seasonal movement from electronic tagging."}

# Tag and release information
release_sites <- tagging %>%
  distinct(tag_id, release_lon, release_lat, release_region)

# Recapture / detection events
detections <- tagging %>%
  filter(!is.na(recapture_lon)) %>%
  mutate(season = case_when(
    lubridate::month(as.Date(date)) %in% c(12, 1, 2) ~ "Winter",
    lubridate::month(as.Date(date)) %in% c(3, 4, 5)  ~ "Spring",
    lubridate::month(as.Date(date)) %in% c(6, 7, 8)  ~ "Summer",
    TRUE                                               ~ "Autumn"
  ))

# Displacement distance (Haversine)
detections <- detections %>%
  mutate(displacement_km = geosphere::distHaversine(
    cbind(release_lon, release_lat),
    cbind(recapture_lon, recapture_lat)
  ) / 1000)

season_colors <- c(Winter = "#4575B4", Spring = "#91CF60",
                   Summer = "#FC8D59", Autumn = "#D73027")

# --- Map of trajectories ---
p5_map <- ggplot() +
  geom_sf(data = europe, fill = "grey85", color = "grey60", linewidth = 0.3) +
  geom_segment(data = detections,
               aes(x = release_lon, y = release_lat,
                   xend = recapture_lon, yend = recapture_lat,
                   color = season),
               alpha = 0.4, linewidth = 0.5,
               arrow = arrow(length = unit(0.08, "inches"), type = "open")) +
  geom_point(data = release_sites,
             aes(x = release_lon, y = release_lat),
             shape = 21, fill = "yellow", color = "black",
             size = 3, stroke = 0.6) +
  scale_color_manual(values = season_colors, name = "Season") +
  coord_sf(xlim = c(-12, 10), ylim = c(38, 56), expand = FALSE) +
  labs(x = "Longitude", y = "Latitude") +
  theme_bw(base_size = 12) +
  theme(panel.background = element_rect(fill = "#D6EAF8"))

# --- Displacement distance by season ---
p5_disp <- ggplot(detections, aes(x = season, y = displacement_km, fill = season)) +
  geom_boxplot(outlier.shape = 16, outlier.size = 1.5, alpha = 0.8) +
  scale_fill_manual(values = season_colors) +
  labs(x = NULL, y = "Displacement distance (km)") +
  theme_bw(base_size = 12) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 30, hjust = 1))

# --- Monthly displacement heatmap (individual × month) ---
monthly_disp <- detections %>%
  mutate(month = lubridate::month(as.Date(date), label = TRUE)) %>%
  group_by(tag_id, month) %>%
  summarise(mean_disp = mean(displacement_km, na.rm = TRUE), .groups = "drop")

p5_heat <- ggplot(monthly_disp, aes(x = month, y = tag_id, fill = mean_disp)) +
  geom_tile() +
  scale_fill_viridis_c(option = "plasma", name = "Mean disp. (km)", na.value = "grey90") +
  labs(x = "Month", y = "Individual tag") +
  theme_minimal(base_size = 10) +
  theme(axis.text.y = element_blank())

figure5 <- (p5_map | (p5_disp / p5_heat)) +
  plot_layout(widths = c(1.8, 1)) +
  plot_annotation(tag_levels = "A")

figure5

ggsave("figures/Figure5_tagging.pdf", figure5,
       width = 14, height = 7, useDingbats = FALSE)
ggsave("figures/Figure5_tagging.png", figure5,
       width = 14, height = 7, dpi = 300)
```

---

## FIGURE 6 ##

Integrated comparison of genomic ancestry proportions and tagging-based connectivity
metrics across sampling sites, illustrating congruence between the two approaches.

```{r figure6, fig.cap="Figure 6. Integrated genomic and tagging comparison."}

# --- Atlantic ancestry proportion per site (from ADMIXTURE K = 2) ---
genomic_summary <- admix_K2_df %>%
  group_by(site, region, longitude) %>%
  summarise(
    mean_atlantic  = mean(V1),
    se_atlantic    = sd(V1) / sqrt(n()),
    .groups = "drop"
  ) %>%
  arrange(longitude)

# --- Connectivity index from tagging (proportion of fish moving > 500 km) ---
tag_connectivity <- detections %>%
  group_by(release_region) %>%
  summarise(
    n_total    = n(),
    n_distant  = sum(displacement_km > 500, na.rm = TRUE),
    connect_idx = n_distant / n_total,
    se_connect  = sqrt(connect_idx * (1 - connect_idx) / n_total),
    .groups = "drop"
  )

# Merge genomic and tagging summaries on shared region
integrated <- genomic_summary %>%
  left_join(tag_connectivity, by = c("region" = "release_region"))

# --- Panel A: genomic ancestry by longitude ---
p6a <- ggplot(genomic_summary,
              aes(x = longitude, y = mean_atlantic, color = region)) +
  geom_errorbar(aes(ymin = mean_atlantic - se_atlantic,
                    ymax = mean_atlantic + se_atlantic),
                width = 0.5, alpha = 0.5) +
  geom_point(size = 3) +
  geom_smooth(aes(group = 1), method = "loess", se = FALSE,
              color = "black", linewidth = 0.8, linetype = "dashed") +
  scale_color_manual(values = region_colors, name = "Region") +
  labs(x = "Longitude (°E)", y = "Atlantic ancestry proportion") +
  theme_bw(base_size = 12)

# --- Panel B: tagging connectivity by longitude ---
p6b <- ggplot(integrated %>% filter(!is.na(connect_idx)),
              aes(x = longitude, y = connect_idx, color = region)) +
  geom_errorbar(aes(ymin = connect_idx - se_connect,
                    ymax = connect_idx + se_connect),
                width = 0.5, alpha = 0.5) +
  geom_point(size = 3) +
  geom_smooth(aes(group = 1), method = "loess", se = FALSE,
              color = "black", linewidth = 0.8, linetype = "dashed") +
  scale_color_manual(values = region_colors, name = "Region") +
  labs(x = "Longitude (°E)", y = "Tagging connectivity index") +
  theme_bw(base_size = 12) +
  theme(legend.position = "none")

# --- Panel C: scatter of genomic vs tagging across sites ---
p6c <- ggplot(integrated, aes(x = mean_atlantic, y = connect_idx, color = region)) +
  geom_errorbar(aes(ymin = connect_idx - se_connect,
                    ymax = connect_idx + se_connect),
                width = 0.01, alpha = 0.5) +
  geom_errorbarh(aes(xmin = mean_atlantic - se_atlantic,
                     xmax = mean_atlantic + se_atlantic),
                 height = 0.01, alpha = 0.5) +
  geom_point(size = 3.5) +
  geom_text_repel(aes(label = site), size = 3, box.padding = 0.4) +
  geom_smooth(aes(group = 1), method = "lm", se = TRUE,
              color = "black", fill = "grey80", linewidth = 0.8) +
  scale_color_manual(values = region_colors, name = "Region") +
  labs(x = "Atlantic ancestry proportion",
       y = "Tagging connectivity index") +
  theme_bw(base_size = 12)

figure6 <- (p6a / p6b | p6c) +
  plot_layout(widths = c(1, 1.2)) +
  plot_annotation(tag_levels = "A")

figure6

ggsave("figures/Figure6_integrated.pdf", figure6,
       width = 14, height = 8, useDingbats = FALSE)
ggsave("figures/Figure6_integrated.png", figure6,
       width = 14, height = 8, dpi = 300)
```
